---
title: "STA 610L: Module 0.0"
subtitle: "Template"
author: "Dr. Olanrewaju Michael Akande"
date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
library(tufte)
library(sjPlot)
library(lme4)
library(sjstats)
library(merTools)
library(lattice)
```



## NELS data recap

We fit a random intercepts ANCOVA to assess the association between math score and SES.

That is, we allowed school-specific intercepts while including SES as a covariate $x_{ij}$:
$$y_{ij} = \beta_{0,j}+\beta_1x_{ij} + \varepsilon_{ij},$$
with
$$
\begin{split}
y_{ij} & = \beta_{0,j} + \beta_1x_{ij} + \varepsilon_{ij}; \ \ \ i = 1,\ldots, n_j\\
\varepsilon_{ij} & \overset{iid}{\sim} N\left(0,\sigma^2\right); \ \ \ \  \beta_{0,j} \sim \mathcal{N} \left( \mu, \tau^2 \right); \ \ \ j = 1, \ldots, J,
\end{split}
$$

This model allows separate intercepts for each school but assumes a common slope.

One concern is whether SES has the same relationship with math scores in all schools.

For example, some schools may have less of a disparity in scores across levels of SES than others.




---
## Random intercepts model


```{r echo=FALSE, cache=TRUE, fig.height=5.2}
load('data/nels.Rdata')
avmscore.schools <- tapply(nels$mscore,nels$school,mean,na.rm=TRUE)
id.schools <- names(avmscore.schools)
m <- length(id.schools)
nels$sesstd <- nels$ses/sd(nels$ses)
m5 <- lmer(mscore~sesstd+(1|school),data=nels)

xplot=seq(-2.9,2.3,by=.1)
yplot=rep(60,length(xplot))
plot(xplot,yplot,type="n",ylim=c(30,70),xlab="Standardized SES",ylab="Math Score")
for(school in 1:length(id.schools))
{
  yplot=coef(m5)$school[school,1]+coef(m5)$school[school,2]*xplot
  lines(xplot,yplot)
}
```



---
## Extending the model

As an initial step towards extending our random intercepts model, we can examine the variation in slopes across 685 separate regression models fit separately in each school:
$$y_{ij}=\beta_{0,j}+\beta_{1,j}x_{ij}+\varepsilon_{ij}, ~~ \varepsilon_{ij} \sim N(0,\sigma^2_j).$$.

Here, we fit an "unpooled" model with completely different regressions for each school.

In each case, the OLS estimaste
$$\widehat{\beta}_j=(X_j'X_j)^{-1}X_j'y_j,$$
where $X_j$ contains a column of 1's for the intercept and a column containing the SES of each student.







---

class: center, middle

# What's next?

### Move on to the readings for the next module!




